---
title: "Coral Reef Cover Vignette"
output: rmarkdown::html_vignette
---
```{r setup}
library(collaborativedatasciencepackage, quietly = TRUE)
```

```{r library, echo = FALSE}
knitr::opts_chunk$set(
  warning = FALSE,
  message = FALSE
)
# load in required packages
library(tidyr)
library(knitr)
library(vegan)
library(BiodiversityR)
library(modelr)
library(tidybayes)
library(ggplot2)
library(tibble)
library(patchwork)
library(ggrepel)
library(dplyr)
library(grid)
```

# Collaborative Data Science Project

## Introduction
In 2025, benthic imagery was collected on SCUBA across three reef zones (lagoon, back reef, and fore reef) of the Mesoamerican barrier reef in Belize, spanning an inshore to offshore shelf gradient. This gradient exhibits increasing wind and wave exposure towards the open ocean, with the fore reef experiencing the highest wave action. Additionally, lagoonal reefs receive higher nutrient inputs from anthropogenic activities, increasing nutrient concentrations and likely increasing water turbidity. Such environmental differences likely structure benthic communities across the three reef zones. 
At each zone, six 50m transects were conducted by taking quadrat photos every 5m, yielding ten photos per transect. Images were then analyzed using CoralNet, an automated benthic cover annotation platform, to quantify percent cover of benthic taxa within each quadrat. With these data, we are interested in examining two questions:

1. ***How does benthic cover vary across three reef zones of the Belize Barrier Reef?***
2. ***How is coral abundance influenced by habitat type and macroalgae abundance?***

#### 1) Metadata

This dataset contains the metadata for quadrat photos taken every 5m along 50m transects. Six transects were conducted at each of the three study sites in Belize: lagoon (whale shoal), fore reef, and back reef. The dataset contains the **image name**, **site of collection**, **date of collection**, **transect number**, and **quadrat number**.

```{r metadata}
data("metadata", package = "collaborativedatasciencepackage")
head(metadata)
```

#### 2) Percent Cover Dataset

This dataset consists of percent cover data for each quadrat photo. Percent cover was calculated using CoralNet, a software that automates the analysis of coral reef benthic images. Essentially, CoralNet generates 25 random points over each image, identifies/labels the points, and then calculates percent cover from that data. The dataset contains the **image ID**, **image name** (same as above), **annotation status** (whether the image has been labeled), **number of collection points**, and various columns corresponding to **different benthic categories** such as hard coral species, soft coral species, macroalgal species, etc. See the [Glossary](#glossary) for a key to each category.

```{r percent-cover-data}
data("per.cover.data", package = "collaborativedatasciencepackage")
head(per.cover.data)
```

## Data Wrangling

Before we begin our analyses, we first must wrangle our data so it's ready for downstream analyses. First, we need to combine the metadata with the percent cover data.

```{r data-join}
benthic.data <- metadata |>
  left_join(per.cover.data, by = "Image.name")
kable(head(benthic.data))
```

Now, we have to calculate the average percent cover for each category at each transect. To do this, we first must `gather()` our data so that we have a column for the categories and a column for the percent cover of each category. At this time, we will also add a new column with the names of the reef zones: lagoon, forereef, backreef.

```{r benthic-data}
benthic.data <- benthic.data |>
  gather(key = category, value = percent_cover, 9:69) |> # gather all category columns into one column, "category"
  mutate(Habitat = case_when(Site == "BZE_CBC_BACKREEF" ~ "Backreef",
                            Site == "BZE_CBC_FOREREEF" ~ "Forereef",
                            Site == "BZE_whaleShoal" ~ "Lagoon"),
          Habitat = factor(Habitat, levels = c("Lagoon", "Backreef", "Forereef"))) # reorder the levels of those habitats to mirror the gradient
```

Next, we can group by Habitat, Transect, and Category to calculate the average percent cover for each category at each transect across the ten quadrats.

```{r category-cover}
category.cover <- benthic.data |>
  group_by(Habitat, Transect, category) |>
  summarize(transect_cover = sum(percent_cover) / 1000, .groups = "drop") # divide by 1000 because there are 10 quadrats in a single transect: 100% * 10
head(category.cover)
```

This dataset now shows the proportion of each category found in each transect. However, we also want to classify the categories into broad functional groups for downstream analyses.

```{r functional-group-classify}
# classify functional groups
func.group.cover <- category.cover |>
  mutate(func_group = case_match(
    category,
    
    # macroalgae
    c("Brown_UFM", "Green_UFM", "Red_UFM", "Dictyota", "Halimeda", "Ulva", "Lobophora", "Sargassum", "Mac_matrix",
      "Brown_Enc", "Red_Enc") ~ "Macroalgae",
    
    # corals
    c("AcrCerv", "AcrPalm", "AgAga", "AgTen", "AgUnd", "BL_OrbFav", "BORB", "CoNat", "DiplLab", "FaviaFrag", "MontaCav", "OrbAnn", "OrbFav", "OrbFrank", "PorAstr", "Por_branch", "PorPor", "PseuCliv", "PseuStrig", "Siderastr", "SidSid") ~ "Coral",
    
    # gorgonians
    c("Gorg_Fan", "Gorg_Plume", "Gorg_Whip", "GORG-BRI") ~ "Gorgonian",
    
    # EAM
    c("EAM_HR", "EAM_LR", "EAM_RBL") ~ "EAM",
    
    # Cyanobacteria
    c("Cyano_HR", 
      "Cyano_LR", "Cyano_RBL", "Cyano_Sand") ~ "Cyanobacteria",
    
    # crustose coralline algae (CCA)
    c("CCA_LR", "CCA_HR", "CCA_RBL") ~ "CCA",
    
    # sponges
    c("Spong_Crus", "Spong_Erec", "Spong_Glo") ~ "Sponge",
    
    # seagrass
    c("SyrFil", "ThalTes") ~ "Seagrass",
    
    # all other categories
    .default = "Other"))


# calculate percent cover for each functional group in each transect
func.group.cover <- func.group.cover |>
  select(-category) |>
  group_by(Habitat, Transect, func_group) |>
  summarise(transect_cover = sum(transect_cover), .groups = "drop") # use `sum()` because different categories were grouped
head(func.group.cover)
```

After calculating the percent cover of each functional group for each transect, we need to format the data into "wide" format using `spread()` for downstream analyses. Thus, each row is an observation and each column is a unique variable.

```{r benthic-wide}
benthic.wide <- func.group.cover |>
  spread(key = func_group, value = transect_cover, fill = 0) # set missing values as "0"
head(benthic.wide)
```
## Data Analysis
#### *1) How does benthic cover vary across three reef zones of the Belize Barrier Reef?*
## Principal Coordinate Analysis
Principal Coordinate Analysis (PCoA) is a technique used to visualize multidimensional data. Since we have benthic cover data (functional group abundances from 0-1), we can use the Manhattan distance measure to calculate pairwise distances between each transect in our dataset. This will, ultimately, form a distance matrix that we can then analyze using PCoA to determine how similar or different the benthic cover of each transect is to one another.\
*Note: The Bray-Curtis distance measure could also be used for abundance data, but here we chose the Manhattan distance to analyze the absolute differences in percent cover, rather than relative metrics.*

First, we need to split the metadata (containing the information of the sites and transects) and the community-matrix (containing the percent cover of each functional group on each transect) into two separate datasets before performing a PCoA.

```{r creating-datasets}
# create metadata
benthic.meta <- benthic.wide |>
  select(Habitat, Transect)

# create percent cover dataset
benthic.base <- benthic.wide |>
  ungroup() |>
  select(-Habitat, -Transect)
```

Next, we need to create our distance matrix. This can be done using the `vegdist()` function which calculates pairwise dissimilarities between pairs of transects. Essentially, this will yield the absolute difference in percent cover across all functional groups — ecologically, it will represent the “total amount of compositional change between two transects.”

```{r benthic-distance}
# distance matrix using the Manhattan distance measure
benthic.dist <- vegdist(benthic.base, method = "manhattan")
```

The `wcmdscale()` function performs a Principal Coordinate Analysis. It takes the distance matrix and finds coordinates in a Euclidean space that best preserve those distances. `k = 2` → projects into 2 dimensions (for plotting). `eig = TRUE` → returns eigenvalues (variance explained per axis). Unlike Principal Component Analysis (PCA) that works directly on Euclidean distances, PCoA accepts any distance matrix (e.g., Manhattan, Bray-Curtis).

```{r benthicpcoanoscores}
benthic.pcoa.noscores <- wcmdscale(benthic.dist, k = 2, eig = TRUE)
summary(benthic.pcoa.noscores)

# extract the percentage of variance explained by each principal coordinate axis
eig.vals <- benthic.pcoa.noscores$eig
var.explained <- eig.vals/sum(eig.vals[eig.vals > 0]) # calculate the percentage explained by each axis
var.explained
```

#### Scores & Convex Hulls
Now, we need to extract the scores, or the coordinates, for two different components of the ordination: 1) the species scores: scores for the functional groups (columns in the distance matrix) and 2) the site scores: scores for the transects (rows in the distance matrix).

```{r benthic-pcoa}
# extract species scores
benthic.pcoa <- BiodiversityR::add.spec.scores(benthic.pcoa.noscores, benthic.base, Rscale = TRUE, scaling = 1, multi = 1) # these are the coordinates for each functional group on the PCoA

# compile the species scores into a tidy dataset
benthic.species.scores <- as.data.frame(benthic.pcoa$cproj) |>
  mutate(species = rownames(.)) 
head(benthic.species.scores) 

# extract site scores
benthic.site.scores <- as_tibble(benthic.pcoa$points, .name_repair = "minimal") |>
  rename(Dim1 = 1, Dim2 = 2) |>
  bind_cols(benthic.meta)
head(benthic.site.scores)

# create convex hulls
benthic.hulls <- benthic.site.scores |>
  group_by(Habitat) |>
  dplyr::slice(chull(Dim1, Dim2))
head(benthic.hulls)
```

#### Plotting the PCoA

Now we have everything to plot our PCoA: species scores, site scores, and convex hulls.
Using the `benthic_plot()` function from our created package, we can easily create the PCoA from our species scores, site scores, and convex hull data. Please see the [Supplementary Material](#Supplementary Material) for the original code used to create this plot.


```{r benthic-plot, eval = FALSE}
benthic_plot(benthic.species.scores, benthic.site.scores, benthic.hulls)
benthic_plot
```

**RESULTS:** The lack of overlap between convex hulls indicates that each reef habitat has a distinct benthic community. The species loadings show that the lagoon has a greater abundance of macroalgae, while the backreef has higher coral abundance.

While there seems to be distinct separation in the community composition of the reef habitats, it is important to statistically validate these visual differences. To do this, we can run a PERMANOVA using the `adonis2()` function that tests whether the separations are statistically significant.
```{r permanova}
permanova <- adonis2(benthic.dist ~ Habitat, data = benthic.meta, permutations = 999)
permanova
```
**Results:** The difference between habitat types is statistically significant (p = 0.001), and habitat explains 54.1% of the variation in community composition between the transects.

#### *2) How is coral abundance influenced by habitat type and macroalgae abundance?*
## Generalized Linear Model
Coral cover is a widely used indicator of reef condition, so we next examined the factors that may structure coral abundance across our transects. Here, we focused on two drivers with well-established roles in shaping coral recruitment and survival: **macroalgae cover** and **reef habitat.** Disturbances that kill corals create space for fast-growing macroalgae to take over and suppress coral recovery. Habitat context also matters, as the natural gradient from lagoon to back reef to fore reef is accompanied by shifts in water quality, wind and wave exposure, nutrient availability, etc.

#### Model preliminaries

Let's first look at plots of coral cover vs. macroalgae cover and coral cover by habitat.

```{r cm-data}
# isolate model variables and convert character variables to factors
cm.data <- benthic.wide |> 
  ungroup() |>
  select(Habitat, Transect, Coral, Macroalgae) |>
  mutate(Habitat = as.factor(Habitat),
         Transect = as.factor(Transect))

# plot coral vs. macroalgae
coral.algae.plot <- cm.data |>
  ggplot(aes(x = Macroalgae, y = Coral)) +
  geom_point(color = "#9970AB", size = 4) +
  theme_bw() +
  xlab("Macroalgae cover (%)") +
  ylab("Coral cover (%)")

# plot coral cover by habitat
coral.habitat.plot <- cm.data |>
  ggplot(aes(x = Habitat, y = Coral, fill = Habitat)) +
  geom_boxplot(alpha = 0.7) + 
  geom_jitter(aes(fill = Habitat), size = 3, shape = 21) +
  scale_color_manual(values = c("#FB9A29", "#CC6677", "#9970AB")) +
  scale_fill_manual(values = c("#FB9A29", "#CC6677", "#9970AB")) +
  ylab("Coral cover (%)") +
  theme_bw() +
  theme(legend.position = "none")

grid.arrange(coral.algae.plot, coral.habitat.plot, ncol = 2)
```

As we can see from these two plots, there appears to be trends in the abundance of coral with macroalgae and reef habitat. So, we can use modeling to determine the significance/size of those trends.

To test whether we will need to use a general linear model or a generalized linear model, we must check the normality of coral abundance, our response variable.

```{r benthic-wide-two}
benthic.wide |>
  ggplot(aes(x= Coral)) +
  geom_histogram() +
  theme_classic() +
  xlab("Coral Abundance") +
  ylab("Count")
```

From the histogram, you can see that coral abundance is **not** normally distributed, so we must model our data using a different distribution. The beta distribution is appropriate for this as our abundance data is continuous from 0-1.

#### Running the model

We will perform a Bayesian Generalized Linear Model using the package `{brms}`. Our model, in this case, will consist of macroalgae cover and habitat as predictor variables and coral cover as our response variable. We also will need to include transect as a random effect to account for quadrats taken within the same transect. Bayesian modeling was chosen over frequentist modeling to test the probability of the reef habitats being different from one another.

Now we know that we need to run a generalized linear model using the argument `family =  beta.`

```{r model-run, eval = FALSE}
#| cache: false

# run the model

# {brms} is not yet compatible with the new R version 4.6.0; for users with this version, they may load in the model output from the associated github
if (file.exists("coral_model.rds")) {
  coral.glm <- readRDS("coral_model.rds")
  
# for users with a working {brms} package
} else {
  coral.glm <- brm(Coral ~ Habitat * Macroalgae,
                   data = cm.data,
                   family = "beta",
                   seed = 690)
}

summary(coral.glm)
```

**Results:** The Rhat values of 1 indicate that the model converged successfully. The model suggests that none of the effects are clearly positive or negative as the credible intervals are very wide and all cross zero. This indicates that there is a lot of uncertainty in the model.


#### Coral \~ Habitat

Let's explore the model predictions for solely the categorical variable, habitat, on coral cover.
```{r habitat-effect}
habitat_effect(cm.data, coral.glm)
```
**Results:** This figure shows the posterior predicted distributions of coral cover for each habitat, with the dots representing the medians and the lines representing the credible intervals. The fore reef appears to have the highest predicted coral cover, followed by the back reef and lagoon. However, the wide credible intervals suggest that we cannot confidently distinguish coral cover between the three habitats. Overall, though, all three habitats show heavily right-skewed distributions, meaning that low coral cover is most probable across all habitats.

#### Coral ~ Macroalgae

Now let's look at solely the effects of macroalgae cover on coral cover.
```{r macroalgae-effect, eval = FALSE}
macroalgae.effect <- cm.data |>
  
  # create a data grid that holds 100 evenly-spaced values of macroalgae cover across the observed range
  data_grid(Macroalgae = seq(
    min(cm.data$Macroalgae, na.rm = TRUE),
    max(cm.data$Macroalgae, na.rm = TRUE),
    length.out = 100)) |> 
  
  # fix habitat to a single level so we can visualize the marginal effect of macroalgae on coral cover
  mutate(Habitat = "Backreef") |> 
  
  # add predicted coral cover values from the model
  add_epred_draws(coral.glm, 
                  regex = T, 
                  ndraws = 500, # takes 500 posterior draws of coral cover
                  seed = 690) |>

  # plot the predicted coral cover values against macroalgae cover
  ggplot() +
  geom_line(aes(x = Macroalgae, y = .epred, group = .draw), 
            color = "#9970AB",   
            alpha = 1,
            linewidth = 0.1) +
  
  # plot the raw data points
  geom_point(data = cm.data, aes(x = Macroalgae, y = Coral), color = "#9970AB", fill = "#D3C0DD", shape = 21, size = 3) +
  
  # aesthetics
  theme_bw() +
  coord_cartesian(ylim = c(0, 0.2)) +
  xlab("Macroalgae cover (%)") +
  ylab("Predicted coral cover (%)") +
  theme(plot.margin = unit(c(0,0,0,0), 'lines'))

macroalgae.effect
```
**Results:** This figure shows the posterior draws of the relationship between macroalgae cover and predicted coral cover, with each line representing one predicted relationship and the dots representing the observed data points. Overall, there is a general trend that as macroalgae cover increases, coral cover decreases (as most of the draws exhibit an inverse relationship). However, the wide spread of the lines reflects high uncertainty in this relationship.

## Conclusions
The three reef zones examined in this study differ in their benthic community composition, such that the lagoon is dominated by seagrass and macroalgae whereas the back reef is characterized by coral, epilithic algal matrix (EAM), and cyanobacteria. Despite these clear community differences, the model detected no measurable effects of habitat, macroalgae cover, or their interaction on coral cover. This likely reflects the limited statistical power of our small sample size (n = 18). Therefore, while benthic assemblages vary greatly across reef zones in Belize, a larger dataset may be needed to reliably evaluate the influence of habitat and macroalgae on coral cover across reef zones in Belize.

## Acknowledgements
This project draws on data generated through research conducted by Juan Pablo Lozano-Peña and Megan C. Do in November 2025.


## Glossary {#glossary}

-   *Acropora cervicornis* (AcrCerv)
-   *Acropora palmata* (AcrPalm)
-   *Agaricia agaricites* (AgAga)
-   *Agaricia tenuifolia* (AgTen)
-   *Agaricia undata* (AgUnd)
-   ##(BL_OrbFav)##
-   ##(BORB)##
-   *Colpophyllia natans* (CoNat)
-   *Diploria labyrinthiformis* (DiplLab)
-   *Favia fragum* (FaviaFrag)
-   *Montastraea cavernosa* (MontaCav)
-   *Orbicella annularis* (OrbAnn)
-   *Orbicella faveolata* (OrbFav)
-   *Orbicella franksi* (OrbFrank)
-   *Porites astreoides* (PorAstr)
-   ##(Por_branch)##
-   *Porites porites* (PorPor)
-   *Pseudodiploria clivosa* (PseuCliv)
-   *Pseudodiploria strigosa* (PseuStrig)
-   ##(Siderastr)##
-   *Siderastrea siderea* (SidSid)
-   sea anemones (ANEM)
-   feather duster worms (Feat_dust)
-   *Briareum* (GORG-BRI)
-   sea fans (Gorg_Fan)
-   sea plumes (Gorg_Plume)
-   sea whips (Gorg_Whip)
-   hydra (Hydra)
-   fire corals (Millepora)
-   ##(PCAR)##
-   encrusting sponges (Sponge_Crus)
-   ##(Sponge_Erec)##
-   ##(Sponge_Glo)##
-   ##(Film_Sand)##
-   sand (Sand)
-   bare rock (Bare_Rock)
-   ##(CCA_LR)##
-   ##(Cyano_HR)##
-   ##(Cyano_LR)##
-   ##(Cyano_RBL)##
-   ##(Cyano_Sand)##
-   debris (Debris)
-   ##(ACA)##
-   brown encrusting algae (Brown_Enc)
-   brown algae ultra fine matrix (Brown_UFM)
-   ##(CCA_HR)##
-   ##(CCA_RBL)##
-   *Dictyota* sp. (Dictyota)
-   ##(EAM_HR)##
-   ##(EAM_LR)##
-   ##(EAM_RBL)##
-   green algae ultra fine matrix (Green_UFM)
-   *Halimeda* sp. (Halimeda)
-   *Lobophora* sp. (Lobophora)
-   macroalgae matrix (Mac_matrix)
-   red encrusting algae (Red_Enc)
-   red algae ultra fine matrix (Red_UFM)
-   *Sargassum* sp. (Sargassum)
-   *Ulva* sp. (Ulva)
-   *Syringodium filiforme* (SyrFil)
-   *Thalassia testudinum* (ThalTes)

## Supplementary Material
```{r Supplemental, eval = FALSE}
benthic.plot <- ggplot() +
  
   # plot species loadings as vectors
   geom_segment(data = benthic.species.scores,
    aes(x = 0, y = 0, xend = Dim1, yend = Dim2),
    arrow = arrow(length = unit(0.2, "cm")),
    color = "grey10",
    linewidth = 0.4) +
  
  # plot ellipses
  stat_ellipse(data = benthic.site.scores,
    aes(Dim1, Dim2, color = Habitat), type = "norm",
    linetype = 2) +
  
  # plot convex hulls
  geom_polygon(data = benthic.hulls,
    aes(Dim1, Dim2, fill = Habitat, group = Habitat),
    alpha = 0.6, color = "grey23",
    linewidth = 0.1) +
  
  # plot transect coordinates
  geom_point(data = benthic.site.scores,
    aes(Dim1, Dim2, fill = Habitat, shape = Habitat),
    size = 3, alpha=0.9) +
  
  # add species labels
  ggrepel::geom_text_repel(
    data = benthic.species.scores,
    aes(Dim1, Dim2, label = species),
    fontface = "bold") +
  
  # aesthetics
  theme(
    axis.text = element_text(color = "black", size = 12),
    text = element_text(color = "black", size = 6)) +
  scale_color_manual(values = c("#FB9A29", "#CC6677", "#9970AB")) + 
  scale_fill_manual(values = c("#FB9A29", "#CC6677", "#9970AB")) +
  scale_shape_manual(values = c(21, 22, 23)) +
  labs(x = "Dimension 1 (46.7%)", y = "Dimension 2 (18.0%)") + # include percentage of variance explained by each dimension from var.explained
  theme_classic()
```

```{r supplemental2, eval = FALSE}
habitat.effect <- cm.data |>
  
  # create a data grid that holds macroalgae fixed at its mean
  # this allows us to isolate the effects of solely habitat on coral cover
  data_grid(Habitat = unique(Habitat),  # use all unique levels of 'Habitat'
            Macroalgae = mean(Macroalgae, na.rm = TRUE)) |>
  
  # add predicted coral cover values from the model
  add_epred_draws(coral.glm, 
                  regex = T, 
                  ndraws = 200, # takes 200 posterior draws of coral cover
                  seed = 690) |>
  
  # plot the predicted coral values by each habitat
  ggplot(aes(x = .epred, y = Habitat, color = Habitat, fill = Habitat)) +
  stat_histinterval(alpha = 0.8, justification = -0.2) + # plots distribution of predicted values
  # aesthetics
  theme_bw() +
  xlab("Predicted coral cover (%)") +
  ylab("Habitat") +
  scale_color_manual(values = c("#FB9A29", "#CC6677", "#9970AB"))+
  scale_fill_manual(values = c("#FB9A29", "#CC6677", "#9970AB"))
```
